\chapter{Isotropic hardening models}
Several flow stress models have been implemented in \Vaango.  These are described in this chapter.

\section{Linear hardening model}
The linear hardening model in \Vaango has the form
\Beq
  \sigma_y(\Ep) = \sigma_0 + K \Ep
\Eeq
where $\sigma_0$ is the initial yield stress, $K$ is a hardening modulus, and
$\Ep$ is the equivalent plastic strain.

The linear hardening model can be invoked using
\lstset{language=XML}
\begin{lstlisting}
<flow_model type="linear">
  <sigma_0> 700.0e6 </sigma_0>
  <K>1.5e6</K>
</flow_model>
\end{lstlisting}

\section{Johnson-Cook model}
The Johnson-Cook (JC) model (\cite{Johnson1983}) has
the following relation for the flow stress ($\sigma_y$) 
\begin{equation}
  \sigma_y(\Ep,\Edot{\Teq},T) = 
  \left[A + B (\Ep)^n\right]\left[1 + C \ln(\Edot{\star})\right]
  \left[1 - (T^*)^m\right]
\end{equation}
where $\Ep$ is the equivalent plastic strain, 
$A$, $B$, $C$, $n$, $m$ are material constants, and
\begin{equation}
  \Edot{^\star} = \cfrac{\Edot{\Teq}}{\Edot{0}}; \quad
  T^* = \cfrac{(T-T_0)}{(T_m-T_0)}\,.
\end{equation}
In the above equations, $\Edot{\Teq}$ is the 
equivalent strain rate, 
$\Edot{0}$ is a reference strain rate, 
$T_0$ is a reference temperature, and $T_m$ is the melt temperature.  
For conditions where $T^\star < 0$, we assume that $m = 1$.

The inputs for this model have the form
\lstset{language=XML}
\begin{lstlisting}
<flow_model type="johnson_cook">
  <A>792.0e6</A>
  <B>510.0e6</B>
  <C>0.014</C>
  <n>0.26</n>
  <m>1.03</m>
  <T_r>298.0</T_r>
  <T_m>1793.0</T_m>
  <epdot_0>1.0</epdot_0>
</flow_model>
\end{lstlisting}

\section{Steinberg-Guinan model}
The Steinberg-Cochran-Guinan-Lund (SCG) model is a semi-empirical model
that was developed by \cite{Steinberg1980} for high strain rate 
situations and extended to low strain rates and bcc materials by
\cite{Steinberg1989}.  The flow stress in this model is given by
\begin{equation}\label{eq:SCGL}
  \sigma_y(\Ep,\Epdoteq,T) = 
   \left[\sigma_a f(\Ep) + \sigma_t (\Epdoteq, T)\right]
   \frac{\mu(p,T)}{\mu_0} 
\end{equation}
where $\sigma_a$ is the athermal component of the flow stress,
$f(\Ep)$ is a function that represents strain hardening,
$\sigma_t$ is the thermally activated component of the flow stress,
$\mu(p,T)$ is the shear modulus, and $\mu_0$ is the shear modulus 
at standard temperature and pressure.  The strain hardening function
has the form
\begin{equation}
  f(\Ep) = [1 + \beta(\Ep + \Epi)]^n ; \quad
  \sigma_a f(\Ep) \le \sigma_{\text{max}}
\end{equation}
where $\beta, n$ are work hardening parameters, and $\Epi$ is the 
initial equivalent plastic strain.  The thermal component $\sigma_t$
is computed using a bisection algorithm from the following equation (based 
on the work of \cite{Hoge1977})
\begin{equation}
  \Epdoteq = \left[\frac{1}{C_1}\exp\left[\frac{2U_k}{k_b~T}
    \left(1 - \frac{\sigma_t}{\sigma_p}\right)^2\right] + 
    \frac{C_2}{\sigma_t}\right]^{-1}; \quad
  \sigma_t \le \sigma_p
\end{equation}
where $2 U_k$ is the energy to form a kink-pair in a dislocation segment
of length $L_d$, $k_b$ is the Boltzmann constant, $\sigma_p$ is the Peierls
stress. The constants $C_1, C_2$ are given by the relations
\begin{equation}
  C_1 := \frac{\rho_d L_d a b^2 \nu}{2 w^2}; \quad
  C_2 := \frac{D}{\rho_d b^2}
\end{equation}
where $\rho_d$ is the dislocation density, $L_d$ is the length of a 
dislocation segment, $a$ is the distance between Peierls valleys, 
$b$ is the magnitude of the Burgers' vector, $\nu$ is the Debye frequency,
$w$ is the width of a kink loop, and $D$ is the drag coefficient.

The inputs for this model are of the form
\lstset{language=XML}
\begin{lstlisting}
  <flow_model type="steinberg_cochran_guinan">
    <mu_0> 81.8e9 </mu_0>
    <sigma_0> 1.15e9 </sigma_0>
    <Y_max> 0.25e9 </Y_max>
    <beta> 2.0 </beta>
    <n> 0.50 </n>
    <A> 20.6e-12 </A>
    <B> 0.16e-3 </B>
    <T_m0> 2310.0 </T_m0>
    <Gamma_0> 3.0 </Gamma_0>
    <a> 1.67 </a>
    <epsilon_p0> 0.0 </epsilon_p0>
  </flow_model>
\end{lstlisting}

\section{Zerilli-Armstrong model}
The Zerilli-Armstrong (ZA) model (\cite{Zerilli1987,Zerilli1993,Zerilli2004}) 
is based on simplified dislocation mechanics.  The general form of the
equation for the flow stress is
\begin{equation}
  \sigma_y(\Ep,\Epdoteq,T) = 
    \sigma_a + B\exp(-\beta(\Epdoteq) T) + 
                         B_0\sqrt{\Ep}\exp(-\alpha(\Epdoteq) T)
\end{equation}
where $\sigma_a$ is the athermal component of the flow stress given by
\begin{equation}
  \sigma_a := \sigma_g + \frac{k_h}{\sqrt{l}} + K(\Ep)^n,
\end{equation}
$\sigma_g$ is the contribution due to solutes and initial dislocation
density, $k_h$ is the microstructural stress intensity, $l$ is the 
average grain diameter, $K$ is zero for fcc materials, 
$B, B_0$ are material constants.  The functional forms of the exponents 
$\alpha$ and $\beta$ are 
\begin{equation}
  \alpha = \alpha_0 - \alpha_1 \ln(\Epdoteq); \quad
  \beta = \beta_0 - \beta_1 \ln(\Epdoteq); 
\end{equation}
where $\alpha_0, \alpha_1, \beta_0, \beta_1$ are material parameters that
depend on the type of material (fcc, bcc, hcp, alloys).  The Zerilli-Armstrong
model has been modified by \cite{Abed2005} for better performance at high 
temperatures.  However, we have not used the modified equations in our
computations.

The input for this model is of the form
\lstset{language=XML}
\begin{lstlisting}
  <flow_model type="zerilli_armstrong">
    <sigma_g>     46.5e6   </sigma_g>
    <k_H>         5.0e6    </k_H>
    <sqrt_l_inv>  3.7      </sqrt_l_inv>
    <B>           0.0      </B>
    <beta_0>      0.0      </beta_0>
    <beta_1>      0.0      </beta_1>
    <B_0>         890.0e6  </B_0>
    <alpha_0>     0.0028   </alpha_0>
    <alpha_1>     0.000115 </alpha_1>
    <K>           0.0      </K>
    <n>           0.0      </n>
  </flow_model>
\end{lstlisting}

\section{Polymer Zerilli-Armstrong model}
The Zerilli-Armstrong model for polymers has the form:
\begin{equation}
  \sigma_y(\Ep,\Epdoteq,T) = \sigma_g + B\exp(-\beta T^\star) + 
                             B_0\sqrt{\omega \Ep}\exp(-\alpha T^\star)
\end{equation}
where $\sigma_g$ is the athermal component of the flow stress and
\Beq
  \omega = \omega_a + \omega_b \ln(\Epdoteq) + \omega_p \sqrt{\pbar}
\Eeq
where $\omega_a$, $\omega_b$, $\omega_p$ are material parameters
and $\pbar = -p$ is the pressure (positive in compression).
The functional forms of the exponents $\alpha$ and $\beta$ are 
\begin{equation}
  \alpha = \alpha_0 - \alpha_1 \ln(\Epdoteq); \quad
  \beta = \beta_0 - \beta_1 \ln(\Epdoteq); 
\end{equation}
where $\alpha_0, \alpha_1, \beta_0, \beta_1$ are material parameters.
The factors $B$ and $B_0$ are defined as
\Beq
  B = B_{\text{pa}} \left(1 + B_\text{pb} \sqrt{\pbar}\right)^{B_\text{pn}} ~,~~
  B_0 = B_{\text{0pa}} \left(1 + B_\text{0pb} \sqrt{\pbar}\right)^{B_\text{0pn}} 
\Eeq
where $B_\text{pa}$, $B_\text{0pa}$, $B_\text{pb}$, $B_\text{0pb}$,
$B_\text{pn}$, and $B_\text{0pn}$ are material parameters.  Also,
\Beq
  T^\star = \frac{T}{T_0}
\Eeq
where $T_0$ is a reference temperature.

The input tags for the polymer ZA model are:
\lstset{language=XML}
\begin{lstlisting}
  <flow_model type="zerilli_armstrong_polymer">
    <sigma_g>     46.5e6   </sigma_g>
    <B_pa>        0.0      </B_pa>
    <B_pb>        0.0      </B_pb>
    <B_pn>        1.0      </B_pn>
    <beta_0>      0.0      </beta_0>
    <beta_1>      0.0      </beta_1>
    <T_0>         300.0    </T_0>
    <B_0pa>       890.0e6  </B_0pa>
    <B_0pb>       0.0      </B_0pb>
    <B_0pn>       1.0      </B_0pn>
    <omega_a>     0.0      </omega_a>
    <omega_b>     0.0      </omega_b>
    <omega_p>     0.0      </omega_p>
  </flow_model>
\end{lstlisting}

\section{Mechanical thresold stress model}
The Mechanical Threshold Stress (MTS) model 
(\cite{Follans1988,Goto2000a,Kocks2001})  
gives the following form for the flow stress
\begin{equation}
  \sigma_y(\Ep,\Epdoteq,T) = 
    \sigma_a + (S_i \sigma_i + S_e \sigma_e)\frac{\mu(p,T)}{\mu_0} 
\end{equation}
where $\sigma_a$ is the athermal component of mechanical threshold stress,
$\mu_0$ is the shear modulus at 0 K and ambient pressure, 
$\sigma_i$ is the component of the flow stress due to intrinsic barriers 
to thermally activated dislocation motion and dislocation-dislocation 
interactions, $\sigma_e$ is the component of the flow stress due to 
microstructural evolution with increasing deformation (strain hardening), 
($S_i, S_e$) are temperature and strain rate dependent scaling factors.  The
scaling factors take the Arrhenius form
\begin{align}
  S_i & = \left[1 - \left(\frac{k_b~T}{g_{0i}b^3\mu(p,T)}
  \ln\frac{\Epdot{0i}}{\Epdoteq}\right)^{1/q_i}
  \right]^{1/p_i} \\
  S_e & = \left[1 - \left(\frac{k_b~T}{g_{0e}b^3\mu(p,T)}
  \ln\frac{\Epdot{0e}}{\Epdoteq}\right)^{1/q_e}
  \right]^{1/p_e}
\end{align}
where $k_b$ is the Boltzmann constant, $b$ is the magnitude of the Burgers' 
vector, ($g_{0i}, g_{0e}$) are normalized activation energies, 
($\Epdot{0i}, \Epdot{0e}$) are constant reference strain rates, and
($q_i, p_i, q_e, p_e$) are constants.  The strain hardening component
of the mechanical threshold stress ($\sigma_e$) is given by a
modified Voce law
\begin{equation}\label{eq:MTSsige}
  \frac{d\sigma_e}{d\Ep} = \theta(\sigma_e)
\end{equation}
where
\begin{align}
  \theta(\sigma_e) & = 
     \theta_0 [ 1 - F(\sigma_e)] + \theta_{IV} F(\sigma_e) \\
  \theta_0 & = a_0 + a_1 \ln \Epdoteq + a_2 \sqrt{\Epdoteq} - a_3 T \\
  F(\sigma_e) & = 
    \cfrac{\tanh\left(\alpha \cfrac{\sigma_e}{\sigma_{es}}\right)}
    {\tanh(\alpha)}\\
  \ln(\cfrac{\sigma_{es}}{\sigma_{0es}}) & =
  \left(\frac{kT}{g_{0es} b^3 \mu(p,T)}\right)
  \ln\left(\cfrac{\Epdoteq}{\Epdot{0es}}\right)
\end{align}
and $\theta_0$ is the hardening due to dislocation accumulation, 
$\theta_{IV}$ is the contribution due to stage-IV hardening,
($a_0, a_1, a_2, a_3, \alpha$) are constants,
$\sigma_{es}$ is the stress at zero strain hardening rate, 
$\sigma_{0es}$ is the saturation threshold stress for deformation at 0 K,
$g_{0es}$ is a constant, and $\Epdot{0es}$ is the maximum strain rate.  Note
that the maximum strain rate is usually limited to about $10^7$/s.

The inputs for this model are of the form
\lstset{language=XML}
\begin{lstlisting}
  <flow_model type="mechanical_threshold_stress">
    <sigma_a>363.7e6</sigma_a>
    <mu_0>28.0e9</mu_0>
    <D>4.50e9</D>
    <T_0>294</T_0>
    <koverbcubed>0.823e6</koverbcubed>
    <g_0i>0.0</g_0i>
    <g_0e>0.71</g_0e>
    <edot_0i>0.0</edot_0i>
    <edot_0e>2.79e9</edot_0e>
    <p_i>0.0</p_i>
    <q_i>0.0</q_i>
    <p_e>1.0</p_e>
    <q_e>2.0</q_e>
    <sigma_i>0.0</sigma_i>
    <a_0>211.8e6</a_0>
    <a_1>0.0</a_1>
    <a_2>0.0</a_2>
    <a_3>0.0</a_3>
    <theta_IV>0.0</theta_IV>
    <alpha>2</alpha>
    <edot_es0>3.42e8</edot_es0>
    <g_0es>0.15</g_0es>
    <sigma_es0>1679.3e6</sigma_es0>
  </flow_model>
\end{lstlisting}

\section{Preston-Tonks-Wallace model}
The Preston-Tonks-Wallace (PTW) model (\cite{Preston2003}) attempts to 
provide a model for the flow stress for extreme strain rates 
(up to $10^{11}$/s) and temperatures up to melt.  The flow stress is
given by
\begin{equation}
  \sigma_y(\Ep,\Epdoteq,T) = 
     \begin{cases}
       2\left[\tau_s + \alpha\ln\left[1 - \varphi
        \exp\left(-\beta-\cfrac{\theta\Ep}{\alpha\varphi}\right)\right]\right]
       \mu(p,T) & \text{thermal regime} \\
       2\tau_s\mu(p,T) & \text{shock regime}
     \end{cases}
\end{equation}
with 
\begin{equation}
  \alpha := \frac{s_0 - \tau_y}{d}; \quad
  \beta := \frac{\tau_s - \tau_y}{\alpha}; \quad
  \varphi := \exp(\beta) - 1
\end{equation}
where $\tau_s$ is a normalized work-hardening saturation stress,
$s_0$ is the value of $\tau_s$ at 0K,
$\tau_y$ is a normalized yield stress, $\theta$ is the hardening constant
in the Voce hardening law, and $d$ is a dimensionless material
parameter that modifies the Voce hardening law.  The saturation stress
and the yield stress are given by
\begin{align}
  \tau_s & = \max\left\{s_0 - (s_0 - s_{\infty})
     \erf\left[\kappa
       \That\ln\left(\cfrac{\gamma\Xidot}{\Epdoteq}\right)\right],
     s_0\left(\cfrac{\Epdoteq}{\gamma\Xidot}\right)^{s_1}\right\} \\
  \tau_y & = \max\left\{y_0 - (y_0 - y_{\infty})
     \erf\left[\kappa
       \That\ln\left(\cfrac{\gamma\Xidot}{\Epdoteq}\right)\right],
     \min\left\{
       y_1\left(\cfrac{\Epdoteq}{\gamma\Xidot}\right)^{y_2}, 
       s_0\left(\cfrac{\Epdoteq}{\gamma\Xidot}\right)^{s_1}\right\}\right\} 
\end{align}
where $s_{\infty}$ is the value of $\tau_s$ close to the melt temperature,
($y_0, y_{\infty}$) are the values of $\tau_y$ at 0K and close to melt,
respectively, $(\kappa, \gamma)$ are material constants, $\That = T/T_m$,
($s_1, y_1, y_2$) are material parameters for the high strain rate
regime, and
\begin{equation}
  \Xidot = \frac{1}{2}\left(\cfrac{4\pi\rho}{3M}\right)^{1/3}
           \left(\cfrac{\mu(p,T)}{\rho}\right)^{1/2}
\end{equation}
where $\rho$ is the density, and $M$ is the atomic mass.

The inputs for this model are of the form
\lstset{language=XML}
\begin{lstlisting}
  <flow_model type="preston_tonks_wallace">
    <theta> 0.025 </theta>
    <p> 2.0 </p>
    <s0> 0.0085 </s0>
    <sinf> 0.00055 </sinf>
    <kappa> 0.11 </kappa>
    <gamma> 0.00001 </gamma>
    <y0> 0.0001 </y0>
    <yinf> 0.0001 </yinf>
    <y1> 0.094 </y1>
    <y2> 0.575 </y2>
    <beta> 0.25 </beta>
    <M> 63.54 </M>
    <G0> 518e8 </G0>
    <alpha> 0.20 </alpha>
    <alphap> 0.20 </alphap>
  </flow_model>
\end{lstlisting}

\section{SUVIC-I model}
\Textmag{SUVIC-I} is a viscoplastic model for ice that has been used for modeling
the interaction of tires with ice~\cite{Dasari2006,Lee2008}.  The model is an extension of 
the SUVIC model (Strain-rate history dependent Unified Viscoplastic model with
Internal Variables for Crystalline materials).  The model is applicable for strain
rates in the range of $10^{-8}$ to $10^{-2}$ s$^{-1}$.  Further details on the range
of applicability of the model can be found in~\cite{Dasari2006}.

This model is driven by a specialized viscoplastic constitutive model for which
the yield criterion is given by
\Beq
  \left<\frac{X_{ae}-R}{K}\right> ~,~~ \quad 
  \left<x\right> = \begin{cases}
                   x & \quad \text{for} \quad x > 0 \\
                   0 & \quad \text{for} \quad x \le 0 
                   \end{cases}
\Eeq
where the effective reduced stress is defined as
\Beq
  X_{ae} := \sqrt{ \tfrac{3}{2} (\BsT-\Bbeta):(\BsT-\Bbeta)} 
\Eeq
and $R$ is the yield stress, $K$ is the drag stress, $\BsT = \Dev(\Bsig)$ is the deviatoric
stress, $\Bbeta$ is the deviatoric backstress.

The inelastic strain rate is given by
\Beq
  \dot{\BVeps}^i = \left[A\left<\frac{X_{ae}-R}{K}\right>^N \exp\left(-\frac{Q}{RT}\right)\right] \BnT
  ~,~~ \BnT = \tfrac{3}{2} \left(\frac{\BsT-\Bbeta}{X_{ae}}\right)
\Eeq
where $A$ is a kinetic-law material constants, 
and $Q, R, T$ are the activation energy, Universal gas constant and the absolute
temperature, respectively.

The equivalent inelastic strain rate for the model is defined as
\Beq
  \dot{\Veps}^\Teq_i = \sqrt{\tfrac{2}{3} \dot{\BVeps}^i:\dot{\BVeps}^i}
    = A\left<\frac{X_{ae}-R}{K}\right>^N \exp\left(-\frac{Q}{RT}\right) \,.
\Eeq
The evolution of the deviatoric backstress is given by
\Beq
  \dot{\Bbeta} = \tfrac{2}{3} A_1 \dot{\BVeps}^i - A_1 (\beta_\Tsat)^{-1} \dot{\Veps}^\Teq_i \Bbeta
                 - A_2 (\beta^\Teff_\Tsat)^{q-1} C^{-q} \Bbeta
\Eeq
where $A_1$ is a constant that is fitted to the kinematic hardening and dynanamic recovery curves,
$\beta_\Tsat$ is a saturation value of the backstress, and $A_2$, $q$, $C$ are constants fitted
to static recovery curves.
The isotropic hardening yield stress evolves as
\Beq
  \dot{R} = \frac{A_3}{d_g} \dot{\Veps}^\Teq_i \left(1 - \frac{R}{R_\Tsat}\right) - 
            \frac{A_4}{d_g} \frac{(R - R_\Tsat)^p}{C}
\Eeq
where $A_3$ is a constant, $d_g$ is the grain size, $R_\Tsat$ is the saturation value
of $R$, and $A_4$, $p$, $C$ are constants associated with static recovery.
The drag stress evolves as
\Beq
  \dot{K} = \frac{A_5}{d_g} \dot{\Veps}^\Teq_i \left(1 - \frac{K}{K_\Tsat}\right) -
            \frac{A_6}{d_g} \frac{(K - K_\Tsat)^q}{C}
\Eeq
and has the same form as the yield stress evolution rule.  The static recovery terms are
ignored in the \Vaango implementation.
The saturation value for the effective stress is given by
\Beq
  \sigma^\Teff_\Tsat = \sigma_0\left(\frac{\dot{\Veps}^\Teq_i}{\dot{\Veps}_0}\right)^{1/n}
\Eeq
where $\sigma_0$ and $\dot{\Veps}_0$ are reference values and $n$ is a constant.
Similar relations are assumed for the reference values of the other quantities:
\Beq
  \beta^\Teff_\Tsat = \beta_0\left(\frac{\dot{\Veps}^\Teq_i}{\dot{\Veps}_0}\right)^{1/n} ~;~~
  R_\Tsat = R_0\left(\frac{\dot{\Veps}^\Teq_i}{\dot{\Veps}_0}\right)^{1/n} 
\Eeq
The drag stress saturation value is given by
\Beq
  K_\Tsat = \left[\frac{\dot{\Veps}^\Teq_i}{A\exp\left(\frac{-Q}{RT}\right)}\right](X_\Tsat - R_\Tsat)
\Eeq
where $X_\Tsat$ is the saturation value of $X_{ae}$.

A typical input deck for the \Textmag{SUVIC-I} model is shown below.
\lstset{language=XML}
\begin{lstlisting}
  <constitutive_model type="visco_plastic">
    <shear_modulus>3.52e9</shear_modulus>
    <bulk_modulus>8.9e9</bulk_modulus>
    <remove_particles> false </remove_particles>
    <zero_stress_upon_failure> false </zero_stress_upon_failure>
    <stability_check type="none">  </stability_check>
    <equation_of_state type="default_hypo">
      <bulk_modulus>8.9e9</bulk_modulus>
    </equation_of_state>
    <viscoplastic_flow_model type="suvic_i">
      <coeff_backstress_evol>75e6</coeff_backstress_evol>
      <exponent_backstress_evol> 1</exponent_backstress_evol>
      <normalizing_backstress> 1e6</normalizing_backstress>
      <coeff_saturation_backstress> 0.1e6</coeff_saturation_backstress>
      <exponent_backstress> 4</exponent_backstress>
      <ref_strainrate>7.794e-08</ref_strainrate>
      <normalizing_inelastic_strainrate>5.0e9 </normalizing_inelastic_strainrate>
      <activation_energy> 67500.0</activation_energy>
      <universal_gas_constant>8.3144 </universal_gas_constant>
      <temperature>269.15</temperature>
      <exponent_inelastic_strainrate>4.0 </exponent_inelastic_strainrate>
      <coeff_yieldstress_saturation>0.8e6 </coeff_yieldstress_saturation>
      <exponent_yieldstress>4.0 </exponent_yieldstress>
      <coeff_yieldstress_evol> 1600.0e6</coeff_yieldstress_evol>
      <exponent_dragstress> 4.0</exponent_dragstress>
      <coeff_dragstress_evol>95e6 </coeff_dragstress_evol>
      <coeff_stress_saturation>1.0e6</coeff_stress_saturation>
      <intial_drag>0.05e6</intial_drag>
      <initial_yield> 0.0</initial_yield>
      <integration_parameter_theta> 0.5</integration_parameter_theta>
    </viscoplastic_flow_model>
  </constitutive_model>
\end{lstlisting}

