\chapter{Plasticity}

Most plasticity models in \Vaango are implemented as stand-alone models with
their own elasticity law, yield condition, flow rule, and internal variable
evolution rules.  However, because of the large number of possible combinations
of these, a few metal plasticity models are available that allow the user to
swap out one set of rules for another.
  
\begin{NoteBox}
  The list below is not comprehensive.  Please see the following chapters
  for details of the models actually available in \Vaango.
\end{NoteBox}

The plasticity implementations typically contain the following:
\begin{enumerate}
    \item \Textmag{An elasticity model}:
      \begin{itemize}
        \item Isotropic linear elastic model. 
        \item Anisotropic linear elastic models.
        \item Isotropic nonlinear elastic models.
        \item Anisotropic nonlinear elastic models.
        \item \Textmag{Equation of state} to determine the pressure (or 
              volumetric response), for example, 
          \begin{itemize}
            \item Mie-Gruneisen equation of state.
          \end{itemize}
        \item \Textmag{Deviatoric stress model} to determine the shear
              response.
          \begin{itemize}
            \item Nadal-LePoac shear modulus model
            \item Steinberg-Guinan shear modulus model
          \end{itemize}
      \end{itemize}
    \item \Textmag{A yield condition}:
      \begin{itemize} 
        \item von Mises yield condition.
        \item Drucker-Prager yield condition.
        \item Mohr-Coulomb yield condition.
        \item Gurson-Needleman-Tvergaard yield condition.
      \end{itemize} 
    \item \Textmag{A flow rule}:
      \begin{itemize}
        \item Associated flow.
        \item Non-Associated flow using either a flow potential or
              a material parameter.
      \end{itemize}
    \item \Textmag{Isotropic hardening}:
      \begin{itemize}
        \item Perfect plasticity (no hardening).
        \item Johnson-Cook plasticity.
        \item Mechanical Threshold Stress (MTS) plasticity.
      \end{itemize}
    \item \Textmag{Kinematic hardening}:
      \begin{itemize}
        \item Isotropic backstress.
        \item Deviatoric backstress.
      \end{itemize}
    \item \Textmag{Isotropic hardening internal variable evolution rules}:
      \begin{itemize}
        \item Mechanical threshold stress evolution.
        \item Gurson's porosity evolution law. 
      \end{itemize}
    \item \Textmag{Kinematic hardening internal variable evolution rules}:
      \begin{itemize}
        \item Ziegler-Prager evolution rule.
        \item Armstrong-Frederick evolution rule
      \end{itemize}
    \item \Textmag{Damage evolution rules}:
      \begin{itemize}
        \item Johnson-Cook damage model.
        \item Brittle damage model.
      \end{itemize}
    \item \Textmag{Melting temperature models}
    \item \Textmag{Specific heat models}
    \item \Textmag{Material stability-based localization models}:
      \begin{itemize}
        \item Acoustic tensor.
        \item Drucker stability.
      \end{itemize}
\end{enumerate}
  
The models used by the main plasticity codes in the current implementation 
of \Vaango are described in the following chapters, followed by 
descriptions of the plasticity algorithms themselves.
