\chapter{Load Curves}
It is often more convenient to apply a specified load at the MPM particles.
The load may be a function of time.  Such a load versus time curve is called
a {\bf load curve}.

In Vaango, the load curve infrastructure is available for general use
(and not only for particles).  However, it has been implemented only for
a special case of pressure loading.

We invoke the load curve in the \verb|<MPM>| section of the input file
using  
\lstset{language=XML}
\begin{lstlisting}
  <use_load_curves> true </use_load_curves>i
\end{lstlisting}  
The default value is \verb|<use_load_curves> false </use_load_curves>|.

In Vaango, a load curve infrastructure is implemented in the file
\verb|.../MPM/PhysicalBC/LoadCurve.h|.  This file is essentially a templated
structure that has the following private data
\lstset{language=XML}
\begin{lstlisting}
  // Load curve information 
  std::vector<double> d_time;
  std::vector<T> d_load;
  int d_id;
\end{lstlisting}
The variable \verb|d_id| is the load curve ID, \verb|d_time| is the time,
and \verb|d_load| is the load.  Note that the load can have any form - scalar,
vector, matrix, etc.

In our current implementation, the actual specification of the load curve 
information is in the \verb|<PhysicalBC>| section of the input file.  The
implementation is limited in that it applies only to pressure boundary 
conditions for some special geometries (the implementation is in 
\verb|.../MPM/PhysicalBC/PressureBC.cc|).  However, the load curve template can
be used in other, more general, contexts.

A sample input file specification of a pressure load curve is shown below.
In this case, a pressure is applied to the inside and outside of a cylinder.
The pressure is ramped up from 0 to 1 GPa on the inside and from 0 to 0.1 MPa
on the outside over a time of 10 microsecs. 
\lstset{language=XML}
\begin{lstlisting}
   <PhysicalBC>
     <MPM>
       <pressure>
         <geom_object>
           <cylinder label = "inner cylinder">
             <bottom>           [0.0,0.0,0.0]   </bottom>
             <top>              [0.0,0.0,.02]   </top>
             <radius>           0.5             </radius>
           </cylinder>
         </geom_object>
         <load_curve>
           <id>1</id>
           <time_point>
             <time> 0 </time>
             <load> 0 </load>
           </time_point>
           <time_point>
             <time> 1.0e-5 </time>
             <load> 1.0e9 </load>
           </time_point>
         </load_curve>
       </pressure>
       <pressure>
         <geom_object>
           <cylinder label = "outer cylinder">
             <bottom>           [0.0,0.0,0.0]   </bottom>
             <top>              [0.0,0.0,.02]   </top>
             <radius>           1.0             </radius>
           </cylinder>
         </geom_object>
         <load_curve>
           <id>2</id>
           <time_point>
             <time> 0 </time>
             <load> 0 </load>
           </time_point>
           <time_point>
             <time> 1.0e-5 </time>
             <load> 101325.0 </load>
           </time_point>
         </load_curve>
       </pressure>
     </MPM>
   </PhysicalBC>
\end{lstlisting}
The complete input file can be found in \verb|Vaango/StandAlone/inputs/MPM/thickCylinderMPM.ups|.


%\end{document}








